The Numerical Accuracy of Computing Electrostatic Interaction in the Inhomogeneous and Correlated Molecular Systems: for the Ewald Summation, SPME and Staggered Mesh Ewald Methods.

### 2012-07-09 longRange.hetero/alltext/publish/resubmit/main.tex
back up

In this paper, we consider a system with periodic boundary condition.
Suppose it is
composed \redc{of} $N$ charged particles located at $\v r_1, \v r_2, \cdots,
\v r_N$ with charges $q_1, q_2, \cdots, q_N$, respectively.
We study the force \redc{on} a testing particle located at $\v r$, which
feels interactions from all charges $q_1, q_2, \cdots, q_N$
in the system, and does not
exert force on them.
The \emph{error force} is defined by the difference between
the exact force and the calculated
force on the testing particle
\cite{wang2012}, and is denoted by $\Delta \v F(\v r)$.
The magnitude of the force computation error 
is defined by the \emph{root mean square (RMS)} \emph{error}, which is
the square root of the second moment of the error force: $\mathcal
E(\v r) = \sqrt{\langle\vert\Delta\v F(\v r)\vert^2\rangle}$.  The
notation $\langle\cdot\rangle$ means the ensemble average.  We also
estimate the the ensemble average of the error force (also called
\emph{mean error force}), namely $\langle\Delta\v F(\v r)\rangle$,
which plays an important role in the error analysis of the
inhomogeneous systems~\cite{wang2012}.

@@

  \\

@@

As an example, we consider a typical three point charge water model
TIP3P~\cite{jorgensen1983comparison}: each hydrogen atom carries a
charge of $q_{\textrm{H}} = +0.417e$ and each oxygen atom carries a
charge of $q_{\textrm{O}} = -0.834e$. Within one molecule, the H-O
bond and H-O-H angle are rigid, so the correlation within one molecule
is considered important.  Therefore, in this case, we define the nearest neighbor by the
atoms within the same water molecule.  Denoting the index set of the oxygen
atoms by $\Omega_{\textrm{O}}$, and the index set of the hydrogen
atoms by $\Omega_{\textrm{H}}$, we have:

@@

@@

  \rangle,\\

@@

% For the second term, do variable change $\v r'' = \v r' + \v s$, we have:
% \begin{align}
%   \textrm{II} = & 
%   q_i^2
%   \int_{\mathbb R^3}\d d\v r'\v K(\v r_i - \v r')
%   \int_{\mathbb R^3}\d d\v s\,
%   \v K(\v r_i - \v r' - \v s)\,
%   \rho_{q^2}(\v r', \v r' + \v s).
% \end{align}
% Split the integral region of the variable $\v s$ into the first
% neighbor shell $\mathcal B(\v r', R_s)$ and the reset $\mathbb R -
% \mathcal B(\v r', R_s)$. In the case of water, the first neighbor
% shell means the region where a single molecule occupies. Within this
% shell, the relative position of the oxygen and two hydrogens are
% always fixed. Outside of the first neighbor shell, the position of
% oxygens and hydrogens are belongs to different molecules, so the their
% positions are assumed to be independent. In most cases, the first
% neighbor shell is small compared to the simulation box size,
% therefore,
% \begin{align}\nonumber
%   \textrm{II}_{\textrm{out}} = & \,
%   q_i^2
%   \int_{\mathbb R^3}\d d\v r'\v K(\v r_i - \v r')
%   \int_{\mathbb R^3 - \mathcal B(\v r', s)}\d d\v s\,
%   \v K(\v r_i - \v r' - \v s)\,
%   \rho_{q^2}(\v r', \v r' + \v s) \\ \nonumber
%   = &\,
%   q_i^2
%   \int_{\mathbb R^3}\d d\v r'\v K(\v r_i - \v r') \rho_q(\v r')
%   \int_{\mathbb R^3 - \mathcal B(\v r', s)}\d d\v s\,
%   \v K(\v r_i - \v r' - \v s)\,
%   \rho_q(\v r' + \v s) \\
%   \approx &\,
%   q_i^2
%   [\v K\ast\rho_{q}]^2 (\v r_i)
% \end{align}

@@

The shift of the mesh can be effectively treated as shifting all
particles on the opposite direction. Since the system is periodic,
the direction of shift is not important. For convenience, we shift
all particles to the positive direction by half the mesh element size, namely
by a vector $\v s = \sum_\alpha\frac1{2K_\alpha}\v a_\alpha$. From
Eqn.~\eqref{eqn:error-A}, we have
\begin{align}
  \frac12\, [A(\v m, \v r) + A(\v m, \v r+\v s)]
  =&
  \sum_{\alpha=1}^3\sum_{l\neq 0}
  \frac12\,
  Z_{\alpha,l}(\v m)\,
  \big[
  (1 + e^{\pi i l})e^{2\pi i l K_\alpha\v a^\ast_\alpha\cdot \v r} - 2
  \big].
\end{align}
Therefore, when $l$ is odd, $1 + e^{\pi i l}$ vanishes. So the $\frac12\,[A(\v
m, \v r) + A(\v m, \v r+\v s)]$ can be approximated by the leading
order:
\begin{align}
  \frac12\, [A(\v m, \v r) + A(\v m, \v r+\v s)]
  \approx&
  -\sum_{\alpha=1}^3\sum_{l\neq 0}
  Z_{\alpha,l}(\v m).
\end{align}
Then from Eqn.~\eqref{eqn:ik.error.force}, we have the error force of
the staggered ik-differentiation:
\begin{align}\label{eqn:ik-st-error-force}
  \Delta\v F^{\textrm{ik,st}}_{\textrm{rec}}(\v r)
  \approx -2q
  \sum_jq_j
  \sum_\alpha\sum_{l\neq 0}
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  g(\v m) Z_{\alpha,l}(\v m)\,e^{2\pi i\v m\cdot(\v r-\v r_j)}.
\end{align}
Therefore, the error force kernel is
\begin{align}
  \v K^{\textrm{ik,st}}_{\textrm{rec}}(\v r)
  = -2
  \sum_\alpha\sum_{l\neq 0}\v G_{\alpha,l}(\v r),
\end{align}
then the corollary~\ref{thm:tmp2} is applicable.
% Moreover, since both the exact and the error
% force are anti-symmetric,  the Newton's third law is preserved by the
% calculated force.
We have the error estimates:
\begin{align}\label{eqn:ik.st.meanf}
  \big\langle
  \Delta\v F_{\textrm{rec}}^{\textrm{ik,st}}(\v r)
  \big\rangle
  & \approx
  - 2 q
  \big[
  \big(
  \sum_{\alpha}\sum_{l\neq 0}\v G_{\alpha,l}
  \big)
  \ast\rho_q\,
  \big] (\v r),
\end{align}
% Also neglect the high wave number terms,
and the RMS force error:
\begin{align}\label{eqn:ik.st.error}
  \big\langle
  \vert\Delta\v F_{\textrm{rec}}^{\textrm{ik,st}}(\v r)\vert^2
  \big\rangle
  \approx&\,
  4q^2\,
  \big[
  \big(
  \sum_{\alpha} \sum_{l\neq 0}  
  \v G_{\alpha,l}
  \big)^2
  \ast \rho_{q^2}\,
  \big] (\v r) +
  \big\langle
  \Delta\v F_{\textrm{rec}}^{\textrm{ik,st}}(\v r)
  \big\rangle^2.
\end{align}
The superscript ``st'' denotes the staggered mesh method.
The mean error force of the staggered ik-differentiation is the
same as the original ik-differentiation.
The RMS error is improved, because
$\big\langle\vert\Delta\v F_{\textrm{rec}}^{\textrm{ik}}(\v r)\vert^2\big\rangle -
\big\langle\vert\Delta\v F_{\textrm{rec}}^{\textrm{ik,st}}(\v r)\vert^2\big\rangle=
2q^2
\big[\,
\big(
\sum_{\alpha} \sum_{l\neq 0}
\vert \v G_{\alpha,l}\vert^2
\big)
\ast \rho_{q^2}
\,\big] (\v r) \geq 0$.
The nearest neighbor approximation to the correlation error is:
\begin{align}\nonumber
  \mathcal E^{\textrm{ik,st}}_{\textrm{correlation}}(\v r)
  =\,&
  4q^2\,
  \big[
  \big(
  \sum_{\alpha} \sum_{l\neq 0}  
  \v G_{\alpha,l}
  \big)
  \cdot
  \big(
  \sum_{\alpha} \sum_{l\neq 0}  
  [\,\hat T_{\textrm{O}}\hat{\v G}_{\alpha,l}]^{\vee}
  \big)
  \ast \rho_{\textrm{O}}\,
  \big] (\v r) \\\nonumber
  \,&+
  4q^2\,
  \big[
  \big(
  \sum_{\alpha} \sum_{l\neq 0}  
  \v G_{\alpha,l}
  \big)
  \cdot
  \big(
  \sum_{\alpha} \sum_{l\neq 0}  
  [\,\hat T_{\textrm{H}}\hat{\v G}_{\alpha,l}]^{\vee}
  \big)
  \ast \rho_{\textrm{H}}\,
  \big] (\v r).
\end{align}

For the analytical differentiation, similarly we have: $\frac12\,[\v
B(\v m,\v r) + \v B(\v m, \v r+\v s)] \approx 0$. Therefore, from
Eqn.~\eqref{eqn:ana.error.force}, the error force of the staggered
analytical differentiation is approximately the same as the staggered
ik-differentiation, so are all the error estimates:
\begin{align}\label{eqn:ana-st-error-force}
  \Delta\v F^{\textrm{ana,st}}_{\textrm{rec}}(\v r)
  &\approx
  \Delta\v F^{\textrm{ik,st}}_{\textrm{rec}}(\v r)\\
  \big\langle
  \Delta\v F_{\textrm{rec}}^{\textrm{ana,st}}(\v r)
  \big\rangle
  &\approx
  \big\langle
  \Delta\v F_{\textrm{rec}}^{\textrm{ik,st}}(\v r)
  \big\rangle\\
  \big\langle
  \vert\Delta\v F_{\textrm{rec}}^{\textrm{ana,st}}(\v r)\vert^2
  \big\rangle
  &\approx
  \big\langle
  \vert\Delta\v F_{\textrm{rec}}^{\textrm{ik,st}}(\v r)\vert^2
  \big\rangle\\
  \mathcal E^{\textrm{ik,st}}_{\textrm{correlation}}(\v r)
  &\approx
  \mathcal E^{\textrm{ana,st}}_{\textrm{correlation}}(\v r)
\end{align}
The correction of the self-interacting error is not necessary,
because it is automatically canceled by the staggered mesh.
% The error from the self-interaction term is automatically fixed by the
% staggered mesh.
The RMS force error is improved because it is not
difficult to see 
$\big\langle\vert\Delta\v F_{\textrm{rec}}^{\textrm{ana}}(\v r)\vert^2\big\rangle -
\big\langle\vert\Delta\v F_{\textrm{rec}}^{\textrm{ana,st}}(\v r)\vert^2\big\rangle
\geq 0$.
Eqn.~\eqref{eqn:ana-st-error-force} implies, to the
leading order of approximation, the staggered
mesh ik- and analytical differentiation
are equivalent. Considering the latter uses
only one half of  FFTs as the former, it might be preferable
in large scale parallel simulations.
From Eqn.~\eqref{eqn:ik-st-error-force}
and~\eqref{eqn:ana-st-error-force}, it is obvious that the Newton's
third law is better preserved by the staggered mesh methods.

@@

\begin{figure}
  \centering
  \includegraphics[]{rand1.rho.eps}
  \caption{
    The charge number density  of Example 1.
    The red line: positive charge. The blue line: negative charge.
    Since the charges are uniformly distributed
    on $y$ and $z$ direction, the densities are averaged on $y$ and $z$,
    and plotted as a
    function of $x$.
  }
  \label{fig:tmp-rho1}
\end{figure}

In this example,  41472 charges are put into a
$14.90\textsf{nm}\times 7.45\textsf{nm}\times 7.45\textsf{nm}$
periodic simulation box. One third of
them are carrying negative charge of $-0.834e$, and the other two thirds
are carrying positive charge of $+0.417e$.
The position of these charges
are randomly generated subjecting to the number density shown in
Fig.~\ref{fig:tmp-rho1}.  The number density of the  positive
charge   is twice as large as that of the negative charge,
so the
positive and negative charges cancel, and the
system is \emph{locally} neutral. Notice the amount of charges
and the density distributions are intentionally chosen the same as
Example 3~(see section \ref{sec:example3}), 
for an easy comparison.  The only difference is that the positions
of charges are independent  in this example,
while they are correlated in Example 3.
% Both the density distributions are highly
% inhomogeneous, take the positive distribution for example: between 11
% \textsf{nm} and 29 \textsf{nm}, the charge density is
% $1.0\;\textsf{e/nm}^3$, while between 0 \textsf{nm} to 9 \textsf{nm}
% and between 31 \textsf{nm} to 40 \textsf{nm}, the density is
% $0.05\;\textsf{e/nm}^3$.

@@

\begin{figure}
  \centering
  % \includegraphics[width=.48\textwidth]{fig.new/rand2/fig.rho.eps}
  \includegraphics[]{rand2.rho.eps}
  \caption{Charge distribution of Example 2. The red line
    denotes the positive charge distribution, while the blue line
    denotes the negative charge distribution. Since all distributions
    are uniform on $y$ and $z$ direction, they are plotted as a
    function of $x$.}
  \label{fig:tmp-rho2}
\end{figure}

In this example, the setting is nearly the same as
Example 1. The only difference is that all positive charges
are moved to the region where the density of negative charge is low,
see Fig.~\ref{fig:tmp-rho2}.
In this system, the negative charges are separated from their
counter ions, but the whole system is kept neutral. This case rarely
happens in real simulation, but it serves as a good test of the error
estimates in extreme situations.

@@

This example studies the gas-liquid phase equilibrium of a water
system. The size of the simulation box and number of charges are the
same as Example 1 (Sec.~\ref{sec:example1}): 13824 TIP3P water
molecules~\cite{jorgensen1983comparison} are put into a
$14.90\textsf{nm}\times 7.45\textsf{nm}\times 7.45\textsf{nm}$
periodic simulation box.  The molecules dynamics simulations of this
system was performed by GROMACS 4.5~\cite{hess2008gromacs}.  The system
is coupled to a velocity rescale thermostat~\cite{bussi2007canonical}
at temperature 300~\textsf{K}, and simulated long enough to reach the
equilibrium. After the equilibrium,
the water molecules separate into the liquid and vapor phases.
The number densities of oxygen and hydrogen atoms
are the same as those shown by Fig.~\ref{fig:tmp-rho1}.
50 consequential snapshots of
the system are taken along the MD trajectory
with a time interval of 20~\textsf{ps}.  These
snapshots are used to calculate the real error and the charge densities
for the error estimates.

\begin{figure}
  \centering
  \includegraphics[]{fig.water.orig.error.eps}
  \caption{
    Example 3: the real RMS errors and the corresponding
    estimates of original ik- and analytical differentiation.
    Red: the direct part.
    Green: the original ik-differentiation.
    Blue: the original analytical differentiation.
    ``$\times$'': the real error.
    Dashed line: the error estimate, without the correlation error estimate.
    Solid  line: the error estimate, with the nearest neighbor
    approximation of the correlation error.
    The working parameters are the same as Example 1.
    The black arrows indicate the improvement
    of error estimate by including nearest neighbor
    approximation of correlation error.
  }   
  \label{fig:water-error0}
\end{figure}

>>>>

In this paper, we consider a system with periodic boundary condition.
Suppose it is
composed \redc{of} $N$ charged particles located at $\v r_1, \v r_2, \cdots,
\v r_N$ with charges $q_1, q_2, \cdots, q_N$, respectively.
We study the force \redc{on} a testing particle located at $\v r$, which
feels interactions from all charges $q_1, q_2, \cdots, q_N$
in the system, and does not
exert force on them.
% \recheck{
% Here we implicitly assumes that the testing particle is independent
% with the system, but it is not the case in real systems. For example,
% in the water system considered in Sec.~\ref{sec:tmp3}, each atom is bonded
% to two neighboring atoms (via the rigid H-O bond and H-O-H angle).
% Fortunately, this may not be a big problem, firstly because the bonded pair
% electrostatic interacions are considert to be \emph{not} calculated;
% secondly because the non-bonded pair correlation contributes only
% minority part to the total error (see also Sec.~\ref{sec:tmp3}).
% }
The \emph{error force} is defined by the difference between
the exact force and the calculated
force on the testing particle
\cite{wang2012}, and is denoted by $\Delta \v F(\v r)$.
The magnitude of the force computation error 
is defined by the \emph{root mean square (RMS)} \emph{error}, which is
the square root of the second moment of the error force: $\mathcal
E(\v r) = \sqrt{\langle\vert\Delta\v F(\v r)\vert^2\rangle}$.  The
notation $\langle\cdot\rangle$ means the ensemble average.  We also
estimate the the ensemble average of the error force (also called
\emph{mean error force}), namely $\langle\Delta\v F(\v r)\rangle$,
which plays an important role in the error analysis of the
inhomogeneous systems~\cite{wang2012}.

@@

  \\\label{eqn:start-corr}

@@

As an example, we consider a typical three point charge water model
TIP3P~\cite{jorgensen1983comparison}: each hydrogen atom carries a
charge of $q_{\textrm{H}} = +0.417e$ and each oxygen atom carries a
charge of $q_{\textrm{O}} = -0.834e$.  \recheck{ There are two types
  of correlation in the system: the bonded correlation and the
  non-bonded correlation. The former originates from the rigid H-O
  bond and H-O-H angle within one molecule. The latter originates from
  the intermolecular non-bonded interaction: the van der Waals and
  the electrostatic interaction.
}
% Within one molecule, the H-O bond and
% H-O-H angle are rigid, so the correlation within one molecule is
% considered important.
% Therefore, in this case,

We firstly consider the bonded correlation, and define the nearest
neighbor by the atoms within the same water molecule.  Denoting the
index set of the oxygen atoms by $\Omega_{\textrm{O}}$, and the index
set of the hydrogen atoms by $\Omega_{\textrm{H}}$, we have:

@@

\begin{align}\label{eqn:tom}

@@

  \rangle,\\\label{eqn:thm}

@@

\recheck{
It sould be noted that we define the force error by inserting a
testing charge into the water system, and comparing the
calculated and exact force on it. This definition implicitly assumes
the independency between the testing particle and the
system. Therefore, the correlation error
estimate~\eqref{eqn:c-error-water}
actually neglects the correlation
of the testing particle to the system, and
calculates the error from all bonded correlations
within its {neighboring} molecules, as illustrated by Fig.~\ref{fig:tmp-add0}.
This is consistent
with the usual setting of the bonded atoms: the electrostatic interaction
between them is considered to be \emph{not} calculated.
Throughout this paper, we also do not calculated this interaction.
% the intramolecular non-bonded
% interaction is not calculated.
}

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{correlation.1.eps}
  \caption{The schematic plot of two interacting water molecules.
    When we calculate the force error of the left oxygen atom, the
    correlation with its bonded hydrogens are not considered, but
    the correlations within any neighboring molecule are considered,
    as indicated by the solid lines with two arrows.}
  \label{fig:tmp-add0}
\end{figure}

\recheck{ If the system is homogeneous and isotropic, the pair
  correlation function $\rho^{(2)}(\v r', \v r'')$ in
  Eqn.~\eqref{eqn:start-corr} can be represented by the radial
  distribution functions (RDFs).
  Following the same discussion as the bonded correlation,
  the non-bonded correlation error can also be estimated.
  The functions $T_{\textrm{O}}$ and $T_{\textrm{H}}$
  are given by
  \begin{align}\label{eqn:nb-nna-1}
    \hat T_{\textrm{O}} (\v m)
    &= \int_0^{\infty}
    [q_{_{\textrm{H}}} \rho_{_{\textrm{H}}} \tilde g_{_{\textrm{OH}}}(r)
    +q_{_{\textrm{O}}} \rho_{_{\textrm{O}}} \tilde g_{_{\textrm{OO}}}(r)]
    \,\frac{\sin(2\pi m r)}{2\pi mr}\,4\pi r^2 \d dr,\\\label{eqn:nb-nna-2}
    \hat T_{\textrm{H}} (\v m)
    &= \int_0^{\infty}
    [q_{_{\textrm{H}}} \rho_{_{\textrm{H}}} \tilde g_{_{\textrm{HH}}}(r)
    +q_{_{\textrm{O}}} \rho_{_{\textrm{O}}} \tilde g_{_{\textrm{OH}}}(r)]
    \,\frac{\sin(2\pi m r)}{2\pi mr}\,4\pi r^2 \d dr,
  \end{align}
  where
  $\tilde g_{_{\textrm{OO}}}(r) = g_{_{\textrm{OO}}}(r) - 1$,
  $\tilde g_{_{\textrm{OH}}}(r) = g_{_{\textrm{OH}}}(r) - 1$ and
  $\tilde g_{_{\textrm{HH}}}(r) = g_{_{\textrm{HH}}}(r) - 1$.
  $g_{_{\textrm{OO}}}(r)$, $g_{_{\textrm{OH}}}(r)$ and $g_{_{\textrm{HH}}}(r)$
  are the RDFs of oxygen-oxygen, oxgygen-hydrogen and hydrogen-hydrogen,
  respectively. $\rho_{_{\textrm{H}}}$ and $\rho_{_{\textrm{O}}}$ are
  the homogeneous number densities
  of the hydrogen and oxgygen atoms, respectively.
  For a short-range correlated system, the RDFs converges to 1
  as $r$ goes to infinity, so the integrals in Eqn.~\eqref{eqn:nb-nna-1}
  and \eqref{eqn:nb-nna-2} are converging. Practically, we integrate
  them to a upper bond denoted by $R_g$.

  It can be seen that both the bonded and non-bonded correlation error
  estimates require some uniformity of the correlation in the system:
  the bonded estimate requires the uniformity of the water
  direction, while the non-bonded estimate requires the uniformity
  of the density distribution and the isotropicity of the system.
  These restrictions should be carefully investigated before using
  these error estimates.
  If the restrictions are seriously perturbed,
  one can either uses these estimates for 
  some regions (or sub-system of the whole system) that
  still preserve the uniformity,
  or estimate the correlation error at the cost of
  $\mathcal O(N^2\log N)$, if the correlation in the system is highly
  inhomogeneous.  
  % one can either estimate the correlation error at the cost of
  % $\mathcal O(N^2\log N)$, or estimate the correlation error for
  % some regions (or sub-system of the whole system) that
  % still preserve those uniformity.

}
% For the second term, do variable change $\v r'' = \v r' + \v s$, we have:
% \begin{align}
%   \textrm{II} = & 
%   q_i^2
%   \int_{\mathbb R^3}\d d\v r'\v K(\v r_i - \v r')
%   \int_{\mathbb R^3}\d d\v s\,
%   \v K(\v r_i - \v r' - \v s)\,
%   \rho_{q^2}(\v r', \v r' + \v s).
% \end{align}
% Split the integral region of the variable $\v s$ into the first
% neighbor shell $\mathcal B(\v r', R_s)$ and the reset $\mathbb R -
% \mathcal B(\v r', R_s)$. In the case of water, the first neighbor
% shell means the region where a single molecule occupies. Within this
% shell, the relative position of the oxygen and two hydrogens are
% always fixed. Outside of the first neighbor shell, the position of
% oxygens and hydrogens are belongs to different molecules, so the their
% positions are assumed to be independent. In most cases, the first
% neighbor shell is small compared to the simulation box size,
% therefore,
% \begin{align}\nonumber
%   \textrm{II}_{\textrm{out}} = & \,
%   q_i^2
%   \int_{\mathbb R^3}\d d\v r'\v K(\v r_i - \v r')
%   \int_{\mathbb R^3 - \mathcal B(\v r', s)}\d d\v s\,
%   \v K(\v r_i - \v r' - \v s)\,
%   \rho_{q^2}(\v r', \v r' + \v s) \\ \nonumber
%   = &\,
%   q_i^2
%   \int_{\mathbb R^3}\d d\v r'\v K(\v r_i - \v r') \rho_q(\v r')
%   \int_{\mathbb R^3 - \mathcal B(\v r', s)}\d d\v s\,
%   \v K(\v r_i - \v r' - \v s)\,
%   \rho_q(\v r' + \v s) \\
%   \approx &\,
%   q_i^2
%   [\v K\ast\rho_{q}]^2 (\v r_i)
% \end{align}

@@

The shift of the mesh can be effectively treated as shifting all
particles on the opposite direction. Since the system is periodic,
the direction of shift is not important. For convenience, we shift
all particles to the positive direction by half the mesh element size, namely
by a vector $\v s = \sum_\alpha\frac1{2K_\alpha}\v a_\alpha$. From
Eqn.~\eqref{eqn:error-A}, we have
\begin{align}
  \frac12\, [A(\v m, \v r) + A(\v m, \v r+\v s)]
  =&
  \sum_{\alpha=1}^3\sum_{l\neq 0}
  \frac12\,
  Z_{\alpha,l}(\v m)\,
  \big[
  (1 + e^{\pi i l})e^{2\pi i l K_\alpha\v a^\ast_\alpha\cdot \v r} - 2
  \big].
\end{align}
Therefore, when $l$ is odd, $1 + e^{\pi i l}$ vanishes. So the $\frac12\,[A(\v
m, \v r) + A(\v m, \v r+\v s)]$ can be approximated by the leading
order:
\begin{align}\label{eqn:leading-order-1}
  \frac12\, [A(\v m, \v r) + A(\v m, \v r+\v s)]
  \approx&
  -\sum_{\alpha=1}^3\sum_{l\neq 0}
  Z_{\alpha,l}(\v m).
\end{align}
Then from Eqn.~\eqref{eqn:ik.error.force}, we have the error force of
the staggered ik-differentiation:
\begin{align}\label{eqn:ik-st-error-force}
  \Delta\v F^{\textrm{ik,st}}_{\textrm{rec}}(\v r)
  \approx -2q
  \sum_jq_j
  \sum_\alpha\sum_{l\neq 0}
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  g(\v m) Z_{\alpha,l}(\v m)\,e^{2\pi i\v m\cdot(\v r-\v r_j)}.
\end{align}
Therefore, the error force kernel is
\begin{align}
  \v K^{\textrm{ik,st}}_{\textrm{rec}}(\v r)
  = -2
  \sum_\alpha\sum_{l\neq 0}\v G_{\alpha,l}(\v r),
\end{align}
then the corollary~\ref{thm:tmp2} is applicable.
% Moreover, since both the exact and the error
% force are anti-symmetric,  the Newton's third law is preserved by the
% calculated force.
We have the error estimates:
\begin{align}\label{eqn:ik.st.meanf}
  \big\langle
  \Delta\v F_{\textrm{rec}}^{\textrm{ik,st}}(\v r)
  \big\rangle
  & \approx
  - 2 q
  \big[
  \big(
  \sum_{\alpha}\sum_{l\neq 0}\v G_{\alpha,l}
  \big)
  \ast\rho_q\,
  \big] (\v r),
\end{align}
% Also neglect the high wave number terms,
and the RMS force error:
\begin{align}\label{eqn:ik.st.error}
  \big\langle
  \vert\Delta\v F_{\textrm{rec}}^{\textrm{ik,st}}(\v r)\vert^2
  \big\rangle
  \approx&\,
  4q^2\,
  \big[
  \big(
  \sum_{\alpha} \sum_{l\neq 0}  
  \v G_{\alpha,l}
  \big)^2
  \ast \rho_{q^2}\,
  \big] (\v r) +
  \big\langle
  \Delta\v F_{\textrm{rec}}^{\textrm{ik,st}}(\v r)
  \big\rangle^2.
\end{align}
The superscript ``st'' denotes the staggered mesh method.
The mean error force of the staggered ik-differentiation is the
same as the original ik-differentiation.
The RMS error is improved, because
$\big\langle\vert\Delta\v F_{\textrm{rec}}^{\textrm{ik}}(\v r)\vert^2\big\rangle -
\big\langle\vert\Delta\v F_{\textrm{rec}}^{\textrm{ik,st}}(\v r)\vert^2\big\rangle=
2q^2
\big[\,
\big(
\sum_{\alpha} \sum_{l\neq 0}
\vert \v G_{\alpha,l}\vert^2
\big)
\ast \rho_{q^2}
\,\big] (\v r) \geq 0$.
The nearest neighbor approximation to the correlation error is:
\begin{align}\nonumber
  \mathcal E^{\textrm{ik,st}}_{\textrm{correlation}}(\v r)
  =\,&
  4q^2\,
  \big[
  \big(
  \sum_{\alpha} \sum_{l\neq 0}  
  \v G_{\alpha,l}
  \big)
  \cdot
  \big(
  \sum_{\alpha} \sum_{l\neq 0}  
  [\,\hat T_{\textrm{O}}\hat{\v G}_{\alpha,l}]^{\vee}
  \big)
  \ast \rho_{\textrm{O}}\,
  \big] (\v r) \\\nonumber
  \,&+
  4q^2\,
  \big[
  \big(
  \sum_{\alpha} \sum_{l\neq 0}  
  \v G_{\alpha,l}
  \big)
  \cdot
  \big(
  \sum_{\alpha} \sum_{l\neq 0}  
  [\,\hat T_{\textrm{H}}\hat{\v G}_{\alpha,l}]^{\vee}
  \big)
  \ast \rho_{\textrm{H}}\,
  \big] (\v r).
\end{align}

For the analytical differentiation, similarly we have:
\begin{align}\label{eqn:leading-order-2}
\frac12\,[\v
B(\v m,\v r) + \v B(\v m, \v r+\v s)] \approx 0.  
\end{align}
Therefore, from
Eqn.~\eqref{eqn:ana.error.force}, the error force of the staggered
analytical differentiation is approximately the same as the staggered
ik-differentiation, so are all the error estimates:
\begin{align}\label{eqn:ana-st-error-force}
  \Delta\v F^{\textrm{ana,st}}_{\textrm{rec}}(\v r)
  &\approx
  \Delta\v F^{\textrm{ik,st}}_{\textrm{rec}}(\v r)\\
  \big\langle
  \Delta\v F_{\textrm{rec}}^{\textrm{ana,st}}(\v r)
  \big\rangle
  &\approx
  \big\langle
  \Delta\v F_{\textrm{rec}}^{\textrm{ik,st}}(\v r)
  \big\rangle\\
  \big\langle
  \vert\Delta\v F_{\textrm{rec}}^{\textrm{ana,st}}(\v r)\vert^2
  \big\rangle
  &\approx
  \big\langle
  \vert\Delta\v F_{\textrm{rec}}^{\textrm{ik,st}}(\v r)\vert^2
  \big\rangle\\
  \mathcal E^{\textrm{ik,st}}_{\textrm{correlation}}(\v r)
  &\approx
  \mathcal E^{\textrm{ana,st}}_{\textrm{correlation}}(\v r)
\end{align}
The correction of the self-interacting error is not necessary,
because it is automatically canceled by the staggered mesh.
% The error from the self-interaction term is automatically fixed by the
% staggered mesh.
The RMS force error is improved because it is not
difficult to see 
$\big\langle\vert\Delta\v F_{\textrm{rec}}^{\textrm{ana}}(\v r)\vert^2\big\rangle -
\big\langle\vert\Delta\v F_{\textrm{rec}}^{\textrm{ana,st}}(\v r)\vert^2\big\rangle
\geq 0$.
Eqn.~\eqref{eqn:ana-st-error-force} implies, to the
leading order of approximation, the staggered
mesh ik- and analytical differentiation
are equivalent. Considering the latter uses
only one half of  FFTs as the former, it might be preferable
in large scale parallel simulations.
From Eqn.~\eqref{eqn:ik-st-error-force}
and~\eqref{eqn:ana-st-error-force}, it is obvious that the Newton's
third law is better preserved by the staggered mesh methods.

@@

\begin{figure}
  \centering
  \includegraphics[]{fig.rho.eps}
  \caption{
    The charge number density  of Example 1 and 2.
    The red solid line: positive charge of Example 1.
    The red dashed line: positive charge of Example 2.
    The blue line: negative charge.
    Since the charges are uniformly distributed
    on $y$ and $z$ direction, the densities are averaged on $y$ and $z$,
    and plotted as a
    function of $x$.
  }
  \label{fig:tmp-rho}
\end{figure}

In this example,  41472 charges are put into a
$14.90\textsf{nm}\times 7.45\textsf{nm}\times 7.45\textsf{nm}$
periodic simulation box. One third of
them are carrying negative charge of $-0.834e$, and the other two thirds
are carrying positive charge of $+0.417e$.
The position of these charges
are randomly generated subjecting to the number density shown in
Fig.~\ref{fig:tmp-rho}.  The number density of the  positive
charge   is twice as large as that of the negative charge,
so the
positive and negative charges cancel, and the
system is \emph{locally} neutral. Notice the amount of charges
and the density distributions are intentionally chosen the same as
Example 3~(see section \ref{sec:example3}), 
for an easy comparison.  The only difference is that the positions
of charges are independent  in this example,
while they are correlated in Example 3.
% Both the density distributions are highly
% inhomogeneous, take the positive distribution for example: between 11
% \textsf{nm} and 29 \textsf{nm}, the charge density is
% $1.0\;\textsf{e/nm}^3$, while between 0 \textsf{nm} to 9 \textsf{nm}
% and between 31 \textsf{nm} to 40 \textsf{nm}, the density is
% $0.05\;\textsf{e/nm}^3$.

@@

% \begin{figure}
%   \centering
%   % \includegraphics[width=.48\textwidth]{fig.new/rand2/fig.rho.eps}
%   \includegraphics[]{rand2.rho.eps}
%   \caption{Charge distribution of Example 2. The red line
%     denotes the positive charge distribution, while the blue line
%     denotes the negative charge distribution. Since all distributions
%     are uniform on $y$ and $z$ direction, they are plotted as a
%     function of $x$.}
%   \label{fig:tmp-rho2}
% \end{figure}

In this example, the setting is nearly the same as
Example 1. The only difference is that all positive charges
are moved to the region where the density of negative charge is low,
see Fig.~\ref{fig:tmp-rho}.
In this system, the negative charges are separated from their
counter ions, but the whole system is kept neutral. This case rarely
happens in real simulation, but it serves as a good test of the error
estimates in extreme situations.

@@

This example studies the gas-liquid phase equilibrium of a water
system. The size of the simulation box and number of charges are the
same as Example 1 (Sec.~\ref{sec:example1}): 13824 TIP3P water
molecules~\cite{jorgensen1983comparison} are put into a
$14.90\textsf{nm}\times 7.45\textsf{nm}\times 7.45\textsf{nm}$
periodic simulation box.  The molecules dynamics simulations of this
system was performed by GROMACS 4.5~\cite{hess2008gromacs}.  The system
is coupled to a velocity rescale thermostat~\cite{bussi2007canonical}
at temperature 300~\textsf{K}, and simulated long enough to reach the
equilibrium. After the equilibrium,
the water molecules separate into the liquid and vapor phases.
The number densities of oxygen and hydrogen atoms
are the same as Example 1, see Fig.~\ref{fig:tmp-rho}.
50 consequential snapshots of
the system are taken along the MD trajectory
with a time interval of 20~\textsf{ps}.  These
snapshots are used to calculate the real error and the charge densities
for the error estimates.

\begin{figure}
  \centering
  \includegraphics[]{fig.water.orig.error.eps}
  \caption{
    Example 3: the real RMS errors and the corresponding
    estimates of original ik- and analytical differentiation.
    Red: the direct part.
    Green: the original ik-differentiation.
    Blue: the original analytical differentiation.
    ``$\times$'' or ``+'': the real error.
    Dashed line: the error estimate, without the correlation error estimate.
    Solid  line: the error estimate, with the \recheck{bonded} nearest neighbor
    approximation of the correlation error.
    The working parameters are the same as Example 1.
    The black arrows indicate the improvement
    of error estimate by including \recheck{bonded} nearest neighbor
    approximation of correlation error.
  }   
  \label{fig:water-error0}
\end{figure}
